Deployable Probabilistic Programming

### 2019-09-24 paper/paper.tex
camera ready

\documentclass[sigplan,10pt]{acmart}
\usepackage[utf8]{inputenc}
\usepackage{url}
\usepackage{hyperref}
\usepackage{listings}[language=Golang]
\usepackage{GoMono}

@@

\lstset{ % add your own preferences
    frame=single,
    basicstyle=\ttfamily\footnotesize,
    numbers=none,
    numbersep=4pt,
    showstringspaces=false,
    tabsize=4,
    language=Golang % this is it !
}

\acmConference[Onward! 2019]{SPLASH Onward! 2019}{October 20-25, 2019}{Athens, Greece}

% Copyright
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

@@

In line with the guidelines, we have implemented a probabilistic
programming facility for the Go programming language, Infergo.
%(\url{http://infergo.ml/}).
We have chosen Go because Go is a
small but expressive programming language with efficient
implementation that has recently become quite popular for
computation-intensive server-side programming. Infergo is used
in production environment for inference of mission-critical
algorithm parameters.

@@

Just like it is possible to compose library and user-defined
functions into more complex functions, it should be possible to
compose models to build other models~\cite{THS+17,SSW+18}.
Infergo supports model composition organically: a differentiated
model method can be called from another differentiated method,
not necessarily a method of the same model object or type.
Derivatives can propagate through methods of different models
without restriction. Each model is a building block for another
model. For example, if there are two independent
single-parameter models \lstinline{A} and \lstinline{B}
describing the data, then the product model \lstinline{AB} can
be obtained by composition of \lstinline{A} and \lstinline{B}
(Figure~\ref{fig:composition}).
\begin{figure}
  \begin{lstlisting}[framexleftmargin=10pt,numbers=left]
type A struct {Data []float64}
func (model A) Observe(x []float64) {
  ...
}

@@

In Infergo, distributions are just models, and adding and using
a custom distribution is indistinguishable from adding a model
and calling the model's method from another model. By
convention, library distributions, in addition to
\lstinline{Observe}, also have methods \lstinline{Logp} and
\lstinline{Logps} for the scalar and vector version of $\log$
probability density functions, for convenience of use in models.
The distribution library in Infergo is just a model package
which is automatically differentiated like any other model.  New
distributions can be added in any package.
Figure~\ref{fig:exponential-distribution} shows the Infergo
definition of the exponential distribution.
\begin{figure}
  \begin{lstlisting}[framexleftmargin=10pt,numbers=left]
// Exponential distribution.
type expon struct{}

@@

// Logp computes the log pdf of a single
// observation.
func (_ expon) Logp(lambda float64, y float64)
  float64 {
  logl := math.Log(lambda)
  return logl - lambda*y
}

// Logps computes the log pdf of a vector
// of observations.
func (_ expon) Logps(lambda float64, y ...float64)
  float64 {
  ll := 0.
  logl := math.Log(lambda)
  for i := range y {
    ll += logl - lambda*y[i]
  }
  return ll
}
  \end{lstlisting}
  \caption{The exponential distribution in Infergo.}
  \label{fig:exponential-distribution}
\end{figure}

@@

In Infergo, non-determinism can be easily introduced through
data. In the basic example if Section~\ref{sec:basic-example},
the data is encapsulated in the model as a vector, populated
before inference. However, this is not at all required by
Infergo.  Instead of being stored in a Go slice, the data can be
read from a Go channel or network connection and generated by
another goroutine or process.
Figure~\ref{fig:basic-example-channel} shows a modified model
from the basic example with the data gradually read from a
channel rather than passed as a whole in a field of the model
object.
\begin{figure}
  \begin{lstlisting}[framexleftmargin=10pt,numbers=left]
type StreamingModel struct {
  Data chan float64  // data is a channel
  N    int           // batch size
}

@@

The eight schools problem~\cite{GCS+03} is an application of
the hierarchical normal model that often serves as an introductory
example of Bayesian statistical modeling.  In this problem,
experiments were performed to estimate the effects of coaching
programs in each of eight high schools. The outcome variable in
each study was the increase in test score compared to the same
test before the program. The average increase and the standard
deviation of the increase were reported for each school. Based
on that and on assumption of similarity between schools,
expected effects are to be estimated.
Figure~\ref{fig:eight-schools} shows specifications of the same
model in Stan and Infergo, side-by-side.
\begin{figure*}
\begin{minipage}[t]{0.45\textwidth}
  \begin{lstlisting}[language=Stan,framexleftmargin=10pt,numbers=left]
data {
  int<lower=0> J;
  vector[J] y;
  vector<lower=0>[J] sigma;
}

@@

\centering
a. Stan
  \end{minipage}
  \hfill
\begin{minipage}[t]{0.5\textwidth}
\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
type SchoolsModel struct {
  J          int
  Y          []float64
  Sigma      []float64
}

@@

The model in Figure~\ref{fig:linear-regression} specifies linear
regression (approximating linear dependency between input $x$
and output $y$) on unidimensional data. An improper uniform
prior is placed on the parameters $\alpha$,
$\beta$, and $\log \sigma$ (lines 9--10), which are then
conditioned on observations $x, y$ such that
\begin{equation}
y \sim \mathrm{Normal}(\alpha + \beta x, \sigma)
  \label{eqn:linear-regression-1}
\end{equation}
(lines 12--17).
\begin{figure}
\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
type LRModel struct {
  Data  [][]float64
}

@@

// Simulate simulates y based on x and
// regression parameters.
func (m *LRModel) Simulate(x, alpha, beta float64)
    float64 {
  y := alpha + beta*x
  return y
}
\end{lstlisting}
\caption{Linear regression: the simulator is re-used in both
  inference and prediction.}
\label{fig:linear-regression}
\end{figure}
If only the maximum \textit{a posteriori} is estimated, the
parameter values will maximize their likelihood given the
observations:
\begin{equation}
  (\alpha, \beta, \sigma)_{MAP} = \arg \max \prod\limits_i \mathrm{pdf}_\mathrm{Normal}(\alpha + \beta x_i, \sigma)
  \label{eqn:linear-regression-2}
\end{equation}

@@

The Gaussian mixture model is a probabilistic model of a mixture
of components, where each component comes from a Gaussian
distribution. Mixture models are used to represent a population
consisting of different sub-populations, without indication in
the data to which sub-population each observation belongs. The
model variant in Figure~\ref{fig:gaussian-mixture} accepts a
vector of single-dimensional observations and the number of
components, and infers the distribution of parameters of each
component in the mixture. Following Stan User's
Guide~\cite{SDT18}, component memberships are summed out to make
the model differentiable.
\begin{figure}
\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
type GMModel struct {
  Data  []float64 // observations
  NComp int       // number of components
}

@@

// logSumExp computes log(exp(x)+exp(y)) robustly.
func LogSumExp(x, y float64) float64 {
  z := x
  if y > z {
    z = y
  }
  return z + math.Log(math.Exp(x-z)+math.Exp(y-z))
}

@@

@@

\begin{table*}
  \begin{tabular}{r | c |  c | c | c | c | c | }
    {\it model}         & \multicolumn{6}{c|}{\it time, seconds} \\
                  & \multicolumn{3}{c|}{\it compilation} & \multicolumn{3}{c|}{\it execution} \\
                  & Infergo & Turing & Stan              & Infergo  & Turing & Stan    \\\hline
              Eight schools & 0.50    & -      & 50                & 0.60     & 2.8    & 0.12    \\
     Gaussian mixture model & 0.50    & -      & 54                & 32       & 14     & 4.9     \\
Latent Dirichlet allocation & 0.50    & -      & 54                & 8.9      & 12     & 3.7     \\
  \end{tabular}
\caption{Compilation and execution times for 1000 iterations
  of HMC with 10 leapfrog steps.}
\label{tab:memory-runtime}
\end{table*}
It is customary to include performance evaluation on benchmark
problems in publications on new probabilistic programming
frameworks~\cite{PW14,WVM14,TMY+16,GXG18}. We provide here a
comparison of running times of Infergo and two other probabilistic
programming frameworks: Turing~\cite{GXG18} and
Stan~\cite{Stan17}. Turing is a Julia~\cite{Julia14} library
and DSL offering exploratory statistical modeling  and
inference composition. Stan is a compiled language and a library
optimized for high performance.

We use three models of different size and structure for the
comparison: two of the models, Eight schools and Gaussian
mixture model, were introduced as a part of the case
studies~(Section~\ref{sec:case-studies}); the code for the third
model, Latent Dirichlet allocation~\cite{BNG03}, is included in
Appendix~\ref{app:lda}. The code and data for all models were
originally included in Stan examples and translated for use
with Turing and Infergo. \textit{Eight schools} is the smallest model.
One may expect that most of the time execution time is spent in
executing the boilerplate code rather than the model. \textit{Gaussian
mixture model} is a more elaborated model which benefits greatly
from vectorization. The greater the number of observations (1000
data points were used in the evaluation), the more significant
is the benefit of vectorized computations. \textit{Latent
Dirichlet allocation} is, again, a relatively elaborated model,
which is, however, not as easy to vectorize.

@@

\bibliography{refs}

\clearpage
\appendix

\begin{figure*}
\section{Latent Dirichlet Allocation Model}
  \label{app:lda}
\vspace{\baselineskip}

\begin{minipage}[t]{0.49\textwidth}
\centering
\textbf{Infergo}

\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
type LDAModel struct {
  K     int       // num topics
  V     int       // num words
  M     int       // num docs
  N     int       // total word instances
  Word  []int     // word n
  Doc   []int     // doc for word n
  Alpha []float64 // topic prior
  Beta  []float64 // word prior
}

  for (n in 1:N) {
    real gamma[K];
    for (k in 1:K)
      gamma[k] <- log(theta[doc[n],k])
                + log(phi[k,w[n]]);
    increment_log_prob(log_sum_exp(gamma));
  }
}
\end{lstlisting}
\end{minipage}
\end{figure*}

\end{document}

>>>>

\documentclass[sigplan,screen]{acmart}
\usepackage[utf8]{inputenc}
\usepackage{url}
\usepackage{hyperref}
\usepackage{listings}[language=Golang]
\usepackage{GoMono}
\usepackage{flushend}

@@

\lstset{ % add your own preferences
    xrightmargin=1pt,
    frame=single,
    basicstyle=\ttfamily\footnotesize,
    numbers=none,
    numbersep=4pt,
    showstringspaces=false,
    tabsize=4,
    language=Golang % this is it !
}

\setcopyright{acmlicensed}
\acmPrice{15.00}
\acmDOI{10.1145/3359591.3359727}
\acmYear{2019}
\copyrightyear{2019}
\acmISBN{978-1-4503-6995-4/19/10}
\acmConference[Onward! '19]{Proceedings of the 2019 ACM SIGPLAN International Symposium on New Ideas, New Paradigms, and Reflections on Programming and Software}{October 23--24, 2019}{Athens, Greece}
\acmBooktitle{Proceedings of the 2019 ACM SIGPLAN International Symposium on New Ideas, New Paradigms, and Reflections on Programming and Software (Onward! '19), October 23--24, 2019, Athens, Greece}

 \begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10010147.10010257</concept_id>
<concept_desc>Computing methodologies~Machine learning</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10010147.10010341</concept_id>
<concept_desc>Computing methodologies~Modeling and simulation</concept_desc>
<concept_significance>100</concept_significance>
</concept>
<concept>
<concept_id>10010147.10010257.10010293</concept_id>
<concept_desc>Computing methodologies~Machine learning approaches</concept_desc>
<concept_significance>100</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011050.10011017</concept_id>
<concept_desc>Software and its engineering~Domain specific languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011074.10011075</concept_id>
<concept_desc>Software and its engineering~Designing software</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[300]{Computing methodologies~Machine learning}
\ccsdesc[100]{Computing methodologies~Modeling and simulation}
\ccsdesc[100]{Computing methodologies~Machine learning approaches}
\ccsdesc[300]{Software and its engineering~Domain specific languages}
\ccsdesc[500]{Software and its engineering~Designing software}

@@

In line with the guidelines, we have implemented a probabilistic
programming facility for the Go programming language, Infergo
(\url{http://infergo.org/}). We have chosen Go because Go is a
small but expressive programming language with efficient
implementation that has recently become quite popular for
computation-intensive server-side programming. Infergo is used
in production environment for inference of mission-critical
algorithm parameters.

@@

Just like it is possible to compose library and user-defined
functions into more complex functions, it should be possible to
compose models to build other models~\cite{THS+17,SSW+18}.
Infergo supports model composition organically: a differentiated
model method can be called from another differentiated method,
not necessarily a method of the same model object or type.
Derivatives can propagate through methods of different models
without restriction. Each model is a building block for another
model. For example, if there are two independent
single-parameter models \lstinline{A} and \lstinline{B}
describing the data, then the product model \lstinline{AB} can
be obtained by composition of \lstinline{A} and \lstinline{B}
(Figure~\ref{fig:composition}).
\begin{figure}
  \begin{lstlisting}[xleftmargin=11pt,framexleftmargin=10pt,numbers=left]
type A struct {Data []float64}
func (model A) Observe(x []float64) {
  ...
}

@@

In Infergo, distributions are just models, and adding and using
a custom distribution is indistinguishable from adding a model
and calling the model's method from another model. By
convention, library distributions, in addition to
\lstinline{Observe}, also have methods \lstinline{Logp} and
\lstinline{Logps} for the scalar and vector version of $\log$
probability density functions, for convenience of use in models.
The distribution library in Infergo is just a model package
which is automatically differentiated like any other model.  New
distributions can be added in any package.
Figure~\ref{fig:exponential-distribution} shows the Infergo
definition of the exponential distribution.
\begin{figure}
  \begin{lstlisting}[xleftmargin=11pt,framexleftmargin=10pt,numbers=left]
// Exponential distribution.
type expon struct{}

@@

// Logp computes the log pdf of a single
// observation.
func (expon) Logp(lambda float64, y float64)
  float64 {
  logl := math.Log(lambda)
  return logl - lambda*y
}

// Logps computes the log pdf of a vector
// of observations.
func (expon) Logps(lambda float64, y ...float64)
  float64 {
  ll := 0.
  logl := math.Log(lambda)
  for i := range y {
    ll += logl - lambda*y[i]
  }
  return ll
}
  \end{lstlisting}
  \caption{The exponential distribution in Infergo.}
  \label{fig:exponential-distribution}
\end{figure}

@@

In Infergo, non-determinism can be easily introduced through
data. In the basic example if Section~\ref{sec:basic-example},
the data is encapsulated in the model as a vector, populated
before inference. However, this is not at all required by
Infergo.  Instead of being stored in a Go slice, the data can be
read from a Go channel or network connection and generated by
another goroutine or process.
Figure~\ref{fig:basic-example-channel} shows a modified model
from the basic example with the data gradually read from a
channel rather than passed as a whole in a field of the model
object.
\begin{figure}
  \begin{lstlisting}[xleftmargin=11pt,framexleftmargin=10pt,numbers=left]
type StreamingModel struct {
  Data chan float64  // data is a channel
  N    int           // batch size
}

@@

The eight schools problem~\cite{GCS+03} is an application of
the hierarchical normal model that often serves as an introductory
example of Bayesian statistical modeling.  In this problem,
experiments were performed to estimate the effects of coaching
programs in each of eight high schools. The outcome variable in
each study was the increase in test score compared to the same
test before the program. The average increase and the standard
deviation of the increase were reported for each school. Based
on that and on assumption of similarity between schools,
expected effects are to be estimated.
Figure~\ref{fig:eight-schools} shows specifications of the same
model in Stan and Infergo, side-by-side.
\begin{figure*}
\begin{minipage}[t]{0.45\textwidth}
  \begin{lstlisting}[language=Stan,xleftmargin=11pt,framexleftmargin=10pt,numbers=left]
data {
  int<lower=0> J;
  vector[J] y;
  vector<lower=0>[J] sigma;
}

@@

\centering
a. Stan
  \end{minipage}
  \hfill
\begin{minipage}[t]{0.5\textwidth}
\begin{lstlisting}[xleftmargin=11pt,framexleftmargin=10pt,numbers=left]
type SchoolsModel struct {
  J          int
  Y          []float64
  Sigma      []float64
}

@@

The model in Figure~\ref{fig:linear-regression} specifies linear
regression (approximating linear dependency between input $x$
and output $y$) on unidimensional data. An improper uniform
prior is placed on the parameters $\alpha$,
$\beta$, and $\log \sigma$ (lines 9--10), which are then
conditioned on observations $x, y$ such that
\begin{equation}
y \sim \mathrm{Normal}(\alpha + \beta x, \sigma)
  \label{eqn:linear-regression-1}
\end{equation}
(lines 12--17).
\begin{figure}
\begin{lstlisting}[xleftmargin=11pt,framexleftmargin=10pt,numbers=left]
type LRModel struct {
  Data  [][]float64
}

@@

// Simulate simulates y based on x and
// regression parameters.
func (m *LRModel) Simulate(x,alpha,beta float64)
    float64 {
  y := alpha + beta*x
  return y
}
\end{lstlisting}
\caption{Linear regression: the simulator is re-used in both
  inference and prediction.}
\label{fig:linear-regression}
\end{figure}
If only the maximum \textit{a posteriori} is estimated, the
parameter values will maximize their likelihood given the
observations:
\begin{equation}
  (\alpha, \beta, \sigma)_{MAP} = \arg \max \prod\limits_i \mathrm{pdf}_\mathrm{Normal}(\alpha + \beta x_i, \sigma)
  \label{eqn:linear-regression-2}
\end{equation}

@@

The Gaussian mixture model is a probabilistic model of a mixture
of components, where each component comes from a Gaussian
distribution. Mixture models are used to represent a population
consisting of different sub-populations, without indication in
the data to which sub-population each observation belongs. The
model variant in Figure~\ref{fig:gaussian-mixture} accepts a
vector of single-dimensional observations and the number of
components, and infers the distribution of parameters of each
component in the mixture. Following Stan User's
Guide~\cite{SDT18}, component memberships are summed out to make
the model differentiable.
\begin{figure}
\begin{lstlisting}[xleftmargin=11pt,framexleftmargin=10pt,numbers=left]
type GMModel struct {
  Data  []float64 // observations
  NComp int       // number of components
}

@@

// logSumExp computes log(exp(x)+exp(y))
// robustly.
func LogSumExp(x, y float64) float64 {
  z := x
  if y > z {
    z = y
  }
  return z+math.Log(math.Exp(x-z)+math.Exp(y-z))
}

@@

\begin{figure*}
\begin{minipage}[t]{0.49\textwidth}
\centering
\textbf{Infergo}

\begin{lstlisting}[xleftmargin=11pt,framexleftmargin=10pt,numbers=left]
type LDAModel struct {
  K     int       // num topics
  V     int       // num words
  M     int       // num docs
  N     int       // total word instances
  Word  []int     // word n
  Doc   []int     // doc for word n
  Alpha []float64 // topic prior
  Beta  []float64 // word prior
}

  for (n in 1:N) {
    real gamma[K];
    for (k in 1:K)
      gamma[k] <- log(theta[doc[n],k])
                + log(phi[k,w[n]]);
    increment_log_prob(log_sum_exp(gamma));
  }
}
\end{lstlisting}
\end{minipage}
\caption{Implementation of the Latent Dirichlet allocation model in Infergo and Stan.}
\label{fig:lda}
\end{figure*}

@@

\begin{table*}
  \begin{tabular}{r | c |  c | c | c | c | c | }
    {\it model}         & \multicolumn{6}{c|}{\it time, seconds} \\
                  & \multicolumn{3}{c|}{\it compilation} & \multicolumn{3}{c|}{\it execution} \\
                  & Infergo & Turing & Stan              & Infergo  & Turing & Stan    \\\hline
              Eight schools & 0.50    & -      & 50                & 0.60     & 2.8    & 0.12    \\
     Gaussian mixture model & 0.50    & -      & 54                & 32       & 14     & 4.9     \\
Latent Dirichlet allocation & 0.50    & -      & 54                & 8.9      & 12     & 3.7     \\
  \end{tabular}
\caption{Compilation and execution times for 1000 iterations
  of HMC with 10 leapfrog steps.}
\label{tab:memory-runtime}
\end{table*}

It is customary to include performance evaluation on benchmark
problems in publications on new probabilistic programming
frameworks~\cite{PW14,WVM14,TMY+16,GXG18}. We provide here a
comparison of running times of Infergo and two other probabilistic
programming frameworks: Turing~\cite{GXG18} and
Stan~\cite{Stan17}. Turing is a Julia~\cite{Julia14} library
and DSL offering exploratory statistical modeling  and
inference composition. Stan is a compiled language and a library
optimized for high performance.

We use three models of different size and structure for the
comparison: two of the models, Eight schools and Gaussian
mixture model, were introduced as a part of the case
studies~(Section~\ref{sec:case-studies}); the code for the third
model, Latent Dirichlet allocation~\cite{BNG03}, is shown in
Figure~\ref{fig:lda}. The code and data for all models were
originally included in Stan examples and translated for use
with Turing and Infergo. \textit{Eight schools} is the smallest model.
One may expect that most of the time execution time is spent in
executing the boilerplate code rather than the model. \textit{Gaussian
mixture model} is a more elaborated model which benefits greatly
from vectorization. The greater the number of observations (1000
data points were used in the evaluation), the more significant
is the benefit of vectorized computations. \textit{Latent
Dirichlet allocation} is, again, a relatively elaborated model,
which is, however, not as easy to vectorize.

@@

\bibliography{refs}
\end{document}
